---
title: "Asian Options via Monte Carlo Simulation"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

The lecture script `Lecture 13/AsianOption.py` prices arithmetic-average Asian
options under geometric Brownian motion. This vignette shows how to reproduce
and extend that workflow in pure R using the `CompFinanceR` package.

### Learning Goals

- Review the Asian option contract and Monte Carlo pricing setup.
- Simulate path averages with both standard and antithetic sampling.
- Compare estimator precision between sampling methods.

## Package Setup

```{r}
library(CompFinanceR)
```

The package exports GBM specifications and Monte Carlo helpers. We set a seed
for reproducibility throughout the vignette.

```{r}
set.seed(123)
```

## Constructing the Market Model

The GBM parameters match the Python lecture defaults: a 100 spot, 3% drift, and
20% volatility.

```{r}
asian_spec <- gbm_spec(
  initial_value = 100,
  drift = 0.03,
  volatility = 0.2
)
```

## Pricing Calls Across Strikes

The `price_asian_call()` helper wraps the simulation loop and returns tidy
results for any strike vector.

```{r}
call_prices <- price_asian_call(
  process_spec = asian_spec,
  strike = c(90, 100, 110),
  maturity = 1,
  n_paths = 2000,
  n_steps = 252,
  seed = 123
)
call_prices
```

The output includes the standard error estimated from the Monte Carlo sample,
mirroring the summary tables generated in the Python notebook.

## Variance Reduction with Antithetics

`asian_variance_reduction()` evaluates the contract with and without antithetic
variates using a single set of common random numbers.

```{r}
comparison <- asian_variance_reduction(
  process_spec = asian_spec,
  strike = 100,
  maturity = 1,
  option_type = "call",
  n_paths = 5000,
  n_steps = 200,
  seed = 321
)
comparison
```

Antithetic sampling roughly halves the reported standard error, matching the
variance reduction behaviour observed in the original Python material.

## Visualising Strike Sensitivity

A compact plot helps communicate how prices decay with higher strikes.

```{r}
library(ggplot2)

ggplot(call_prices, aes(strike, price)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Asian Call Prices vs Strike",
    x = "Strike",
    y = "Monte Carlo Price"
  )
```

## Key Takeaways

- The new R helpers replicate the lecture outputs with concise syntax.
- Antithetic sampling is opt-in and uses shared random draws for fair
  comparisons.
- Tidy outputs integrate with `dplyr` and `ggplot2` for quick diagnostics.