---
title: "Barrier Options and Hit Probabilities"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

`Lecture 13/DigitalPayoffs_CostReduction.py` introduces barrier contracts and
Monte Carlo diagnostics for hit probabilities. This vignette recreates those
analyses in R with the new barrier utilities.

### Learning Goals

- Price up-and-in and up-and-out calls under GBM dynamics.
- Estimate barrier hit probabilities along simulated paths.
- Illustrate the parity between complemented structures and vanilla options.

## Model and Simulation Inputs

```{r}
library(CompFinanceR)
library(dplyr)
library(plotly)
library(ggplot2)
set.seed(2024)

barrier_spec <- gbm_spec(
  initial_value = 100,
  drift = 0.02,
  volatility = 0.2
)
```

## Pricing a Grid of Up-and-Out Calls

`price_barrier_option()` accepts vectorised strikes and barriers, returning a
tidy tibble with standard errors and hit diagnostics.

```{r}
up_out <- price_barrier_option(
  process_spec = barrier_spec,
  strike = c(95, 100),
  barrier = c(110, 120),
  maturity = 1,
  barrier_type = "up-and-out",
  option_type = "call",
  n_paths = 3000,
  n_steps = 200,
  seed = 2024
)
up_out
```

## Barrier Hit Probability Estimates

The helper `barrier_hit_probability()` isolates the first-passage metric used
in many analytic approximations.

```{r}
up_hits <- barrier_hit_probability(
  process_spec = barrier_spec,
  barrier = c(105, 115, 125),
  maturity = 1,
  barrier_type = "up",
  n_paths = 5000,
  n_steps = 200,
  seed = 2024
)
up_hits
```

## In-Out Parity Check

The Python lecture highlights the identity that an up-and-in call plus an
up-and-out call replicates a vanilla call. We can verify the R implementation
numerically.

```{r}
strike <- 100
barrier <- 125
maturity <- 1

vanilla_price <- price_options(
  black_scholes_spec(
    option_type = "call",
    strike = strike,
    maturity = maturity,
    risk_free_rate = barrier_spec$drift
  ),
  spot = barrier_spec$initial_value,
  volatility = barrier_spec$volatility
)$price

set.seed(303)
out_price <- price_barrier_option(
  process_spec = barrier_spec,
  strike = strike,
  barrier = barrier,
  maturity = maturity,
  barrier_type = "up-and-out",
  option_type = "call",
  n_paths = 8000,
  n_steps = 220,
  seed = 303
)$price

in_price <- price_barrier_option(
  process_spec = barrier_spec,
  strike = strike,
  barrier = barrier,
  maturity = maturity,
  barrier_type = "up-and-in",
  option_type = "call",
  n_paths = 8000,
  n_steps = 220,
  seed = 303
)$price

c(
  vanilla = vanilla_price,
  in_plus_out = in_price + out_price,
  absolute_gap = abs((in_price + out_price) - vanilla_price)
)
```

The residual gap remains within Monte Carlo noise, reproducing the parity
argument from the lecture.

## Visual Summary

```{r}
ggplot(up_out, aes(barrier, price, colour = factor(strike))) +
  geom_point(size = 2) +
  geom_line() +
  labs(
    title = "Up-and-Out Call Prices",
    x = "Barrier",
    y = "Price",
    colour = "Strike"
  )
```

```{r}
plot_ly(up_out, x = ~barrier, y = ~price, color = ~factor(strike),
        type = "scatter", mode = "lines+markers") |>
  layout(
    title = "Up-and-Out Call Prices (Interactive)",
    xaxis = list(title = "Barrier"),
    yaxis = list(title = "Price"),
    legend = list(title = list(text = "Strike"))
  )
```

```{r}
plot_ly(up_hits, x = ~barrier, y = ~hit_probability, type = "scatter",
        mode = "lines+markers", line = list(color = "darkorange")) |>
  layout(
    title = "Barrier Hit Probability",
    xaxis = list(title = "Barrier"),
    yaxis = list(title = "Probability")
  )
```

## Key Takeaways

- `price_barrier_option()` bundles payoff filtering, hit tracking, and
  discounting in a single tidy workflow.
- Hit probabilities are accessible as a standalone diagnostic for stress
  testing.
- In-out parity confirms numerical consistency with the Python teaching
  materials.