---
title: "Simulating Correlated Brownian Motions"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

`Lecture 07/Stochastic Volatility Models/CorrelatedBM.py` constructs correlated
Brownian motions for multi-factor models. The `CompFinanceR` package provides a
compact interface for generating those paths and inspecting the realised
correlations.

### Learning Goals

- Generate correlated Brownian motion increments.
- Verify sample correlations against the target matrix.
- Prepare inputs for multi-factor volatility models.

## Setup

```{r}
library(CompFinanceR)
library(dplyr)
library(ggplot2)
library(plotly)
set.seed(77)
```

## Correlated Brownian Specification

Use `correlated_bm_spec()` to supply the covariance matrix.

```{r}
Sigma <- matrix(c(1, 0.6, 0.6, 1), nrow = 2)
cbm <- correlated_bm_spec(
  initial_values = c(0, 0),
  covariance = Sigma
)

cbm_paths <- simulate_paths(
  cbm,
  n_paths = 5,
  n_steps = 252,
  maturity = 1
)
cbm_paths
```

## Visual Inspection

```{r}
cbm_paths |>
  ggplot(aes(time, value, colour = factor(path_id))) +
  geom_line() +
  facet_wrap(~ component, scales = "free_y") +
  labs(
    title = "Correlated Brownian Motions",
    x = "Time",
    y = "Brownian Level",
    colour = "Path"
  )
```

```{r}
plot_ly(cbm_paths, x = ~time, y = ~value, color = ~factor(path_id),
        type = "scatter", mode = "lines",
        showlegend = FALSE) |>
  layout(
    title = "Correlated Brownian Motions (Interactive)",
    xaxis = list(title = "Time"),
    yaxis = list(title = "Brownian Level")
  )
```

## Realised Correlation

The Python notebook estimates sample correlations between increments. We can
replicate that using tidy operations.

```{r}
realised_corr <- cbm_paths |>
  group_by(path_id) |>
  arrange(time, .by_group = TRUE) |>
  mutate(increment = value - dplyr::lag(value, default = first(value))) |>
  ungroup() |>
  select(path_id, time, component, increment) |>
  tidyr::pivot_wider(names_from = component, values_from = increment) |>
  summarise(
    corr = cor(`component_1`, `component_2`),
    .groups = "drop"
  )
realised_corr
```

```{r}
increment_pairs <- cbm_paths |>
  group_by(path_id, component) |>
  arrange(time, .by_group = TRUE) |>
  mutate(increment = value - dplyr::lag(value, default = first(value))) |>
  ungroup() |>
  select(path_id, time, component, increment) |>
  tidyr::pivot_wider(names_from = component, values_from = increment)

plot_ly(increment_pairs, x = ~`component_1`, y = ~`component_2`,
        color = ~factor(path_id), type = "scatter",
        mode = "markers", marker = list(size = 4, opacity = 0.6),
        showlegend = FALSE) |>
  layout(
    title = "Increment Scatter (Interactive)",
    xaxis = list(title = "ΔW₁"),
    yaxis = list(title = "ΔW₂")
  )
```

The realised correlation is close to the target 0.6, echoing the validation in
the Python script.

## Extended Example: 3 Factors

```{r}
Sigma3 <- matrix(c(
  1, 0.3, -0.2,
  0.3, 1, 0.5,
  -0.2, 0.5, 1
), nrow = 3, byrow = TRUE)
cbm3 <- correlated_bm_spec(initial_values = rep(0, 3), covariance = Sigma3)
paths3 <- simulate_paths(cbm3, n_paths = 1, n_steps = 252, maturity = 1)

tidyr::pivot_wider(paths3, names_from = component, values_from = value) |>
  select(-time, -path_id) |>
  cor()
```

The empirical correlation matrix clusters around the input values, reinforcing
the method highlighted in Lecture 07.

## Key Takeaways

- `correlated_bm_spec()` handles correlation matrices directly and produces tidy
  Brownian motion samples.
- Realised correlations align with the target matrix, matching the verification
  steps in the Python teaching script.
- These samples feed naturally into the stochastic volatility utilities
  implemented elsewhere in the package.