---
title: "Fourier-COS Pricing and Density Recovery"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

`Lecture 08/Fourier Transformation for Option Pricing` demonstrates Fang &
Oosterlee's COS method, density recovery, and FFT-based inversion via the
scripts `CallPut_COS_Method.py`, `COS_Normal_Density_Recovery.py`,
`COS_LogNormal_Density_Recovery.py`, and `DensityRecoveryFFT.py`. The
`CompFinanceR` package now exposes these workflows in R.

### Learning Goals

- Price vanilla calls and puts with the COS method for arbitrary characteristic
  functions.
- Recover densities from characteristic functions using COS and FFT.
- Compare numerical outputs against analytic benchmarks.

## Setup

```{r}
library(CompFinanceR)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
```

## COS Pricing for Black-Scholes

We start with the Black-Scholes characteristic function and call/put pricing,
mirroring `CallPut_COS_Method.py`.

```{r}
spot <- 100
sigma <- 0.2
risk_free <- 0.03
maturity <- 1

cf_bs <- function(u) {
  drift <- log(spot) + (risk_free - 0.5 * sigma^2) * maturity
  exp(1i * u * drift - 0.5 * sigma^2 * maturity * u^2)
}

strikes <- seq(80, 120, by = 5)

cos_calls <- cos_call_put_price(
  cf = cf_bs,
  option_type = "call",
  spot = spot,
  risk_free_rate = risk_free,
  maturity = maturity,
  strikes = strikes
)

cos_puts <- cos_call_put_price(
  cf = cf_bs,
  option_type = "put",
  spot = spot,
  risk_free_rate = risk_free,
  maturity = maturity,
  strikes = strikes
)
```

Compare against analytic Black-Scholes prices to replicate the lecture's
accuracy check.

```{r}
analytic <- tibble::tibble(
  strike = strikes,
  call_bs = price_options(
    black_scholes_spec("call", strike, maturity, risk_free),
    spot = spot,
    volatility = sigma
  )$price,
  put_bs = price_options(
    black_scholes_spec("put", strike, maturity, risk_free),
    spot = spot,
    volatility = sigma
  )$price
)

comparison <- analytic |>
  left_join(cos_calls, by = "strike") |>
  rename(call_cos = price) |>
  left_join(cos_puts, by = "strike") |>
  rename(put_cos = price) |>
  mutate(
    call_error = abs(call_cos - call_bs),
    put_error = abs(put_cos - put_bs)
  )
comparison
```

Errors stay within the tolerance highlighted in the Python notebook.

## Characteristic Function Surfaces

Interactive plots emphasise how the Black-Scholes characteristic function
varies across frequencies and maturities.

```{r}
u_grid <- seq(-40, 40, length.out = 160)
tau_grid <- seq(0.25, 2, length.out = 25)

cf_grid <- expand_grid(
  maturity = tau_grid,
  u = u_grid
) |>
  mutate(
    drift = log(spot) + (risk_free - 0.5 * sigma^2) * maturity,
    cf_value = exp(1i * u * drift - 0.5 * sigma^2 * maturity * u^2),
    real_part = Re(cf_value),
    imag_part = Im(cf_value),
    magnitude = Mod(cf_value)
  )

real_surface <- cf_grid |>
  select(u, maturity, value = real_part) |>
  pivot_wider(names_from = maturity, values_from = value) |>
  arrange(u)

imag_surface <- cf_grid |>
  select(u, maturity, value = imag_part) |>
  pivot_wider(names_from = maturity, values_from = value) |>
  arrange(u)

mag_surface <- cf_grid |>
  select(u, maturity, value = magnitude) |>
  pivot_wider(names_from = maturity, values_from = value) |>
  arrange(u)

u_vals <- real_surface$u
tau_vals <- names(real_surface)[-1] |> as.numeric()
```

```{r}
plot_ly(
  x = tau_vals,
  y = u_vals,
  z = as.matrix(real_surface[, -1]),
  type = "surface",
  colorscale = "RdBu"
) |>
  layout(
    scene = list(
      xaxis = list(title = "Maturity"),
      yaxis = list(title = "Frequency u"),
      zaxis = list(title = "Re(\u03c6(u))")
    ),
    title = "Real Part of the Black-Scholes Characteristic Function"
  )
```

```{r}
plot_ly(
  x = tau_vals,
  y = u_vals,
  z = as.matrix(imag_surface[, -1]),
  type = "surface",
  colorscale = "RdBu"
) |>
  layout(
    scene = list(
      xaxis = list(title = "Maturity"),
      yaxis = list(title = "Frequency u"),
      zaxis = list(title = "Im(\u03c6(u))")
    ),
    title = "Imaginary Part of the Black-Scholes Characteristic Function"
  )
```

```{r}
plot_ly(
  x = tau_vals,
  y = u_vals,
  z = as.matrix(mag_surface[, -1]),
  type = "surface",
  colorscale = "Viridis"
) |>
  layout(
    scene = list(
      xaxis = list(title = "Maturity"),
      yaxis = list(title = "Frequency u"),
      zaxis = list(title = "|\u03c6(u)|")
    ),
    title = "Magnitude of the Black-Scholes Characteristic Function"
  )
```

## Density Recovery with COS

`COS_Normal_Density_Recovery.py` illustrates reconstructing the normal density;
we can recover both normal and log-normal examples using `cos_density_recovery()`.

```{r}
x_grid <- seq(-4, 4, length.out = 501)
normal_cf <- function(u) exp(-0.5 * u^2)

normal_density <- cos_density_recovery(
  cf = normal_cf,
  x = x_grid,
  maturity = 1,
  n_terms = 256,
  truncation = 8
)
```

```{r}
normal_density |>
  ggplot(aes(x, density)) +
  geom_line() +
  labs(title = "Recovered Standard Normal Density")
```

For the log-normal case we compare the recovered density with a kernel density
estimate from Monte Carlo samples, echoing the lecture experiment.

```{r}
log_x <- seq(log(40), log(160), length.out = 501)
log_density <- cos_density_recovery(
  cf = cf_bs,
  x = log_x,
  maturity = maturity,
  n_terms = 256,
  truncation = 8
)

samples <- rlnorm(1e5, meanlog = log(spot) + (risk_free - 0.5 * sigma^2) * maturity, sdlog = sigma * sqrt(maturity))
```

```{r}
log_density |>
  mutate(price = exp(x)) |>
  ggplot(aes(price, density / price)) +
  geom_line(colour = "steelblue") +
  geom_density(data = tibble::tibble(price = samples), aes(price), colour = "orange", inherit.aes = FALSE) +
  labs(title = "Log-Normal Density Recovery")
```

## FFT Density Recovery

`DensityRecoveryFFT.py` applies an FFT to invert characteristic functions. The
`fft_density_recovery()` helper replicates that workflow.

```{r}
fft_density <- fft_density_recovery(
  cf = cf_bs,
  x = log_x,
  maturity = maturity,
  n_terms = 2^12,
  lower_bound = log(40),
  upper_bound = log(160)
)
```

```{r}
fft_density |>
  mutate(price = exp(x)) |>
  ggplot(aes(price, density / price)) +
  geom_line() +
  labs(title = "FFT Density Recovery")
```

## Key Takeaways

- `cos_call_put_price()` and `cos_density_recovery()` deliver the core COS
  routines from Lecture 08 in idiomatic R.
- `fft_density_recovery()` mirrors the FFT inversion script.
- Tidy outputs simplify accuracy checks against analytic benchmarks and Monte
  Carlo estimates.