---
title: "Delta Hedging under Diffusion and Jump Dynamics"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

`Lecture 11/BS_Hedging.py` and `Lecture 11/HedgingWithJumps.py` simulate
discrete-time delta hedging with and without jumps. The new R wrappers expose
the same experiments in a tidy, reproducible pipeline.

### Learning Goals

- Replicate Black-Scholes delta hedging with discrete rebalancing.
- Quantify hedging error dispersion under jump-diffusion dynamics.
- Explore the impact of transaction costs on the cash account.

## Setup

```{r}
library(CompFinanceR)
library(dplyr)
library(plotly)
library(ggplot2)
set.seed(99)
```

## Generate Underlying Paths

We simulate both pure diffusion and Merton jump-diffusion paths on a shared
time grid to mimic the Python lecture setup.

```{r}
paths_gbm <- simulate_paths(
  gbm_spec(initial_value = 100, drift = 0.02, volatility = 0.2),
  n_paths = 256,
  n_steps = 128,
  maturity = 1,
  seed = 99
)

paths_merton <- simulate_paths(
  merton_spec(
    initial_price = 100,
    risk_free_rate = 0.02,
    volatility = 0.2,
    jump_intensity = 0.5,
    jump_mean = -0.1,
    jump_sd = 0.2
  ),
  n_paths = 256,
  n_steps = 128,
  maturity = 1,
  seed = 99
)
```

## Define the Option Contract

```{r}
call_spec <- black_scholes_spec(
  option_type = "call",
  strike = 100,
  maturity = 1,
  risk_free_rate = 0.02
)
```

## Hedging Under GBM

`simulate_delta_hedge_bs()` mirrors the Black-Scholes hedging logic from the
Python lecture.

```{r}
hedge_gbm <- simulate_delta_hedge_bs(
  spot_paths = paths_gbm,
  option_spec = call_spec,
  volatility = 0.2,
  hedge_steps = 128,
  transaction_cost = 0
)

terminal_errors_gbm <- hedge_gbm |>
  dplyr::filter(.data$time == 1) |>
  dplyr::pull(.data$hedging_error)

mean(terminal_errors_gbm)
```

```{r}
plot_ly(x = terminal_errors_gbm, type = "histogram", nbinsx = 60,
        marker = list(color = "steelblue")) |>
  layout(
    title = "Hedging Error Distribution (GBM)",
    xaxis = list(title = "Hedging Error"),
    yaxis = list(title = "Count")
  )
```

## Hedging with Jump Risk

`simulate_delta_hedge_jumps()` delegates to the same engine while accepting
paths produced by jump models.

```{r}
hedge_jump <- simulate_delta_hedge_jumps(
  spot_paths = paths_merton,
  option_spec = call_spec,
  diffusion_volatility = 0.2,
  hedge_steps = 128,
  transaction_cost = 0
)

terminal_errors_jump <- hedge_jump |>
  dplyr::filter(.data$time == 1) |>
  dplyr::pull(.data$hedging_error)

c(
  mean_abs_gbm = mean(abs(terminal_errors_gbm)),
  mean_abs_jump = mean(abs(terminal_errors_jump)),
  sd_gbm = stats::sd(terminal_errors_gbm),
  sd_jump = stats::sd(terminal_errors_jump)
)
```

```{r}
plot_ly() |>
  add_histogram(x = terminal_errors_gbm, name = "GBM", nbinsx = 50,
                opacity = 0.6, marker = list(color = "#1f77b4")) |>
  add_histogram(x = terminal_errors_jump, name = "Jump", nbinsx = 50,
                opacity = 0.6, marker = list(color = "#ff7f0e")) |>
  layout(
    title = "GBM vs Jump Hedging Errors",
    barmode = "overlay",
    xaxis = list(title = "Hedging Error"),
    yaxis = list(title = "Count")
  )
```

Jump risk inflates both the mean absolute error and the standard deviation,
matching the key insight of the lecture.

## Transaction Cost Impact

```{r}
hedge_cost <- simulate_delta_hedge_bs(
  spot_paths = paths_gbm,
  option_spec = call_spec,
  volatility = 0.2,
  hedge_steps = 128,
  transaction_cost = 0.01
)

cash_diff <- hedge_gbm$bond_position - hedge_cost$bond_position

tail(cash_diff, 1)
```

```{r}
plot_ly(hedge_gbm, x = ~time, y = ~bond_position, type = "scatter", mode = "lines",
        name = "No Cost", line = list(color = "forestgreen")) |>
  add_lines(data = hedge_cost, x = ~time, y = ~bond_position,
            name = "With Cost", line = list(color = "firebrick")) |>
  layout(
    title = "Bond Account Trajectories",
    xaxis = list(title = "Time"),
    yaxis = list(title = "Bond Position")
  )
```

The terminal cash shortfall equals the cumulative proportional costs charged on
each rebalancing trade, aligning with the Python implementation.

## Key Takeaways

- Diffusion-only hedging errors concentrate tightly around zero, whereas
  jump-diffusion hedging exhibits fatter tails.
- Transaction costs can be layered onto the same simulation outputs for
  sensitivity analysis.
- The R wrappers retain the pedagogical structure of the Python notebooks while
  providing tidy outputs for downstream summaries.