---
title: "Digital Options with Closed Forms and COS"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

`Lecture 08/CashOrNothing_COS_Method.py` combines analytic Black-Scholes
formulas with Fourier COS density recovery to value cash-or-nothing payoffs.
This vignette demonstrates the equivalent R API in `CompFinanceR`.

### Learning Goals

- Price digital options in closed form under Black-Scholes.
- Apply the COS method to approximate digital prices from a characteristic
  function.
- Validate parity relationships between digital calls and puts.

## Setup

```{r}
library(CompFinanceR)
library(dplyr)
library(plotly)
library(ggplot2)
```

## Closed-Form Pricing

`price_digital_call()` and `price_digital_put()` vectorise over spot and
volatility inputs, returning tidy results ready for plotting.

```{r}
bs_spec <- black_scholes_spec(
  option_type = "call",
  strike = 100,
  maturity = 1,
  risk_free_rate = 0.03,
  dividend_yield = 0.01
)

digital_grid <- price_digital_call(
  model_spec = bs_spec,
  spot = c(95, 100, 105),
  volatility = c(0.18, 0.22)
)
digital_grid
```

## Digital Put-Call Parity

The discounted payout identity from the lecture is preserved in the R helper.

```{r}
call_price <- price_digital_call(
  model_spec = bs_spec,
  spot = 100,
  volatility = 0.2,
  payout = 2
)$price

put_price <- price_digital_put(
  model_spec = bs_spec,
  spot = 100,
  volatility = 0.2,
  payout = 2
)$price

c(
  call = call_price,
  put = put_price,
  discounted_payout = 2 * exp(-0.03)
)
```

## COS Approximation

We can replicate the COS workflow from Python by supplying a characteristic
function. For a log-normal model the analytic density is known, giving us a
simple accuracy check.

```{r}
spot <- 100
sigma <- 0.2
risk_free <- 0.03
maturity <- 1

cf_bs <- function(u) {
  drift <- log(spot) + (risk_free - 0.5 * sigma^2) * maturity
  exp(1i * u * drift - 0.5 * sigma^2 * maturity * u^2)
}

cos_prices <- digital_cos_method(
  cf = cf_bs,
  option_type = "call",
  spot = spot,
  risk_free_rate = risk_free,
  maturity = maturity,
  strikes = c(90, 100, 110),
  n_terms = 256,
  truncation = 8,
  grid_size = 4001
)
cos_prices
```

## Accuracy Diagnostic

```{r}
expected_prices <- exp(-risk_free * maturity) * stats::pnorm(
  (
    log(spot / cos_prices$strike) +
      (risk_free - 0.5 * sigma^2) * maturity
  ) / (sigma * sqrt(maturity))
)

tibble::tibble(
  strike = cos_prices$strike,
  cos_price = cos_prices$price,
  closed_form = expected_prices,
  abs_error = abs(cos_prices$price - expected_prices)
)
```

Errors remain well within the tolerance showcased in the lecture notebook.

## Visual Snapshot

```{r}
ggplot(cos_prices, aes(strike, price)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Digital Call Price vs Strike",
    x = "Strike",
    y = "Price"
  )
```

```{r}
plot_ly(cos_prices, x = ~strike, y = ~price, type = "scatter",
        mode = "lines+markers", line = list(color = "royalblue")) |>
  layout(
    title = "Digital Call Price vs Strike (Interactive)",
    xaxis = list(title = "Strike"),
    yaxis = list(title = "Price")
  )
```

```{r}
plot_ly(digital_grid, x = ~strike, y = ~price, color = ~factor(volatility),
        type = "scatter", mode = "markers", symbol = ~factor(spot)) |>
  layout(
    title = "Closed-Form Digital Prices",
    xaxis = list(title = "Strike"),
    yaxis = list(title = "Price"),
    legend = list(title = list(text = "Volatility"))
  )
```

## Key Takeaways

- Closed-form pricing is exposed directly through `price_digital_call()` and
  `price_digital_put()`.
- The COS density workflow mirrors the Python implementation while benefiting
  from tidy data outputs.
- Put-call parity validation provides a quick confidence check for numerical
  experiments.