---
title: "Forward Start Options under Heston"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

`Lecture 12/HestonForwardStart2.py` prices forward start options by combining a
closed-form characteristic function with the COS method. This vignette walks
through the R translation using `CompFinanceR`.

### Learning Goals

- Review the forward start structure and Heston specification.
- Use the provided characteristic function helper to obtain COS weights.
- Validate results against a Monte Carlo benchmark.

## Heston Specification

```{r}
library(CompFinanceR)
library(dplyr)
library(plotly)
library(ggplot2)

heston_spec <- heston_spec(
  initial_price = 100,
  initial_variance = 0.04,
  risk_free_rate = 0.015,
  dividend_yield = 0,
  mean_reversion = 1.4,
  long_term_variance = 0.04,
  vol_of_vol = 0.3,
  correlation = -0.5,
  scheme = "aes"
)
```

## Pricing with COS

`price_forward_start_heston()` mirrors the Python COS routine while returning a
clean tibble for plotting and analysis.

```{r}
start <- 0.5
maturity <- 1.5
strike_adjustment <- c(0, 0.05, 0.1)

cos_prices <- price_forward_start_heston(
  process_spec = heston_spec,
  start = start,
  maturity = maturity,
  strike_adjustment = strike_adjustment,
  option_type = "call",
  n_terms = 256,
  truncation = 10
)
cos_prices
```

## Monte Carlo Benchmark

A Monte Carlo experiment, echoing the Python script, provides a reasonableness
check on the COS outputs.

```{r}
set.seed(42)
paths <- generate_heston_paths_aes(
  n_paths = 6000,
  n_steps = 360,
  maturity = maturity,
  initial_price = heston_spec$initial_price,
  initial_variance = heston_spec$initial_variance,
  risk_free_rate = heston_spec$risk_free_rate,
  dividend_yield = heston_spec$dividend_yield,
  mean_reversion = heston_spec$mean_reversion,
  long_term_variance = heston_spec$long_term_variance,
  vol_of_vol = heston_spec$vol_of_vol,
  correlation = heston_spec$correlation,
  seed = 42
)

benchmark <- paths |>
  dplyr::group_by(.data$path_id) |>
  dplyr::summarise(
    s_start = stock_price[which.min(abs(time - start))],
    s_maturity = stock_price[which.min(abs(time - maturity))],
    .groups = "drop"
  ) |>
  dplyr::mutate(
    payoff = pmax(s_maturity - (1 + strike_adjustment[1]) * s_start, 0)
  ) |>
  dplyr::summarise(mean_payoff = mean(payoff)) |>
  dplyr::pull(mean_payoff)

adjusted_cos <- cos_prices$price[1]

c(
  cos_price = adjusted_cos,
  mc_price = exp(-heston_spec$risk_free_rate * maturity) * benchmark
)
```

The two pricing approaches align within Monte Carlo error, consistent with the
Python demonstration.

## Strike Sensitivity Plot

```{r}
ggplot(cos_prices, aes(strike_adjustment, price)) +
  geom_point(size = 2) +
  geom_line() +
  labs(
    title = "Forward Start Call Price vs Strike Adjustment",
    x = "Strike Adjustment",
    y = "Price"
  )
```

```{r}
plot_ly(cos_prices, x = ~strike_adjustment, y = ~price,
        type = "scatter", mode = "lines+markers",
        line = list(color = "darkturquoise")) |>
  layout(
    title = "Forward Start Pricing (Interactive)",
    xaxis = list(title = "Strike Adjustment"),
    yaxis = list(title = "Price")
  )
```

```{r}
plot_ly(paths, x = ~time, y = ~stock_price, type = "scatter",
        mode = "lines", line = list(color = "gray"), showlegend = FALSE,
        transforms = list(list(type = "groupby", groups = ~path_id))) |>
  layout(
    title = "Heston Paths Sample",
    xaxis = list(title = "Time"),
    yaxis = list(title = "Stock Price")
  )
```

## Key Takeaways

- `forward_start_characteristic_function()` and
  `price_forward_start_heston()` encapsulate the complex COS workflow.
- Monte Carlo validation remains straightforward via the standard path
  generators.
- Outputs are immediately compatible with tidy plotting utilities.