---
title: "Heston Monte Carlo Simulation"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

`Lecture 10/Monte Carlo Simulation of the Heston Model` focuses on CIR variance
dynamics, discretisation pitfalls, and option pricing accuracy under different
schemes. This vignette rebuilds those experiments with tidy Monte Carlo tools
and interactive Plotly figures.

### Learning Goals

- Inspect CIR variance paths and boundary behaviour.
- Compare Euler and Andersen quadratic-exponential (AES) Heston simulations.
- Benchmark Monte Carlo option prices against COS semi-analytical values.
- Explore the Heston characteristic function across maturities.

## Setup

```{r}
library(CompFinanceR)
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
set.seed(310)
```

## CIR Variance Dynamics

We start with the square-root variance process emphasised in `CIR_paths_boundary.py`.

```{r}
cir_params <- cir_spec(
  initial_value = 0.04,
  mean_reversion = 1.5,
  long_term_mean = 0.04,
  volatility = 0.6
)

cir_paths <- simulate_paths(
  process_spec = cir_params,
  n_paths = 12,
  n_steps = 600,
  maturity = 1,
  seed = 11
)

summary_cir <- cir_paths |>
  summarise(min_variance = min(variance))
summary_cir
```

```{r}
plot_ly(cir_paths, x = ~time, y = ~variance, color = ~factor(path_id),
        type = "scatter", mode = "lines", hoverinfo = "text",
        text = ~paste0("Path ", path_id, "<br>t=", round(time, 3),
                       "<br>v=", round(variance, 4))) |>
  layout(
    title = "CIR Variance Paths (Euler with Reflection)",
    xaxis = list(title = "Time"),
    yaxis = list(title = "Variance"),
    showlegend = FALSE
  )
```

## Heston Specification

We adopt the parameters used in `HestonModelDiscretization.py` (violating the
Feller condition to highlight numerical issues).

```{r}
heston_base <- heston_spec(
  initial_price = 100,
  initial_variance = 0.04,
  risk_free_rate = 0.02,
  dividend_yield = 0,
  mean_reversion = 1.5,
  long_term_variance = 0.04,
  vol_of_vol = 0.9,
  correlation = -0.7,
  scheme = "euler"
)

euler_paths <- simulate_paths(
  heston_base,
  n_paths = 50,
  n_steps = 300,
  maturity = 1,
  seed = 21
)

aes_spec <- heston_spec(
  initial_price = 100,
  initial_variance = 0.04,
  risk_free_rate = 0.02,
  dividend_yield = 0,
  mean_reversion = 1.5,
  long_term_variance = 0.04,
  vol_of_vol = 0.9,
  correlation = -0.7,
  scheme = "aes"
)

aes_paths <- simulate_paths(
  aes_spec,
  n_paths = 50,
  n_steps = 300,
  maturity = 1,
  seed = 21
)
```

### Variance Floor Comparison

```{r}
variance_floor <- bind_rows(
  mutate(euler_paths, scheme = "Euler"),
  mutate(aes_paths, scheme = "AES")
) |>
  group_by(scheme) |>
  summarise(
    min_variance = min(variance),
    mean_terminal_variance = mean(variance[time == max(time)]),
    .groups = "drop"
  )
variance_floor
```

### Interactive Variance Surfaces

```{r}
time_grid <- sort(unique(euler_paths$time))
subset_paths <- euler_paths |>
  filter(path_id <= 20) |>
  select(path_id, time, variance) |>
  pivot_wider(names_from = time, values_from = variance) |>
  arrange(path_id)

subset_paths_aes <- aes_paths |>
  filter(path_id <= 20) |>
  select(path_id, time, variance) |>
  pivot_wider(names_from = time, values_from = variance) |>
  arrange(path_id)

variance_matrix_euler <- as.matrix(subset_paths[, -1])
variance_matrix_aes <- as.matrix(subset_paths_aes[, -1])

plot_ly(x = time_grid, y = subset_paths$path_id, z = variance_matrix_euler,
        type = "surface", colorscale = "Inferno") |>
  layout(
    title = "Variance Surface (Euler)",
    scene = list(
      xaxis = list(title = "Time"),
      yaxis = list(title = "Path"),
      zaxis = list(title = "Variance")
    )
  )
```

```{r}
plot_ly(x = time_grid, y = subset_paths_aes$path_id, z = variance_matrix_aes,
        type = "surface", colorscale = "Viridis") |>
  layout(
    title = "Variance Surface (AES)",
    scene = list(
      xaxis = list(title = "Time"),
      yaxis = list(title = "Path"),
      zaxis = list(title = "Variance")
    )
  )
```

### Stock Price Surfaces

```{r}
stock_surface_euler <- euler_paths |>
  filter(path_id <= 20) |>
  select(path_id, time, stock_price) |>
  pivot_wider(names_from = time, values_from = stock_price) |>
  arrange(path_id)

stock_surface_aes <- aes_paths |>
  filter(path_id <= 20) |>
  select(path_id, time, stock_price) |>
  pivot_wider(names_from = time, values_from = stock_price) |>
  arrange(path_id)

price_matrix_euler <- as.matrix(stock_surface_euler[, -1])
price_matrix_aes <- as.matrix(stock_surface_aes[, -1])

plot_ly(x = time_grid, y = stock_surface_euler$path_id, z = price_matrix_euler,
        type = "surface", colorscale = "PuBuGn") |>
  layout(
    title = "Stock Surface (Euler)",
    scene = list(
      xaxis = list(title = "Time"),
      yaxis = list(title = "Path"),
      zaxis = list(title = "Stock Price")
    )
  )
```

```{r}
plot_ly(x = time_grid, y = stock_surface_aes$path_id, z = price_matrix_aes,
        type = "surface", colorscale = "Cividis") |>
  layout(
    title = "Stock Surface (AES)",
    scene = list(
      xaxis = list(title = "Time"),
      yaxis = list(title = "Path"),
      zaxis = list(title = "Stock Price")
    )
  )
```

## Option Pricing: Euler vs AES vs COS

The lecture compares Monte Carlo schemes against semi-analytic prices. We use
`price_heston_option_cos()` as the reference.

```{r}
strike <- 100
maturity <- 1

terminal_euler <- euler_paths |>
  filter(time == max(time))
terminal_aes <- aes_paths |>
  filter(time == max(time))

discount <- exp(-heston_base$risk_free_rate * maturity)

mc_summary <- tibble::tibble(
  scheme = c("Euler", "AES"),
  price = c(
    discount * mean(pmax(terminal_euler$stock_price - strike, 0)),
    discount * mean(pmax(terminal_aes$stock_price - strike, 0))
  ),
  sd_payoff = c(
    stats::sd(exp(-heston_base$risk_free_rate * maturity) * pmax(terminal_euler$stock_price - strike, 0)),
    stats::sd(exp(-heston_base$risk_free_rate * maturity) * pmax(terminal_aes$stock_price - strike, 0))
  ) / sqrt(nrow(terminal_euler))
)

cos_reference <- price_heston_option_cos(
  process_spec = aes_spec,
  strikes = strike,
  maturity = maturity,
  n_terms = 512,
  truncation = 8
)

mc_summary |>
  mutate(reference = cos_reference$call_price)
```

### Terminal Distribution Comparison

```{r}
combined_terminal <- bind_rows(
  mutate(terminal_euler, scheme = "Euler"),
  mutate(terminal_aes, scheme = "AES")
)

plot_ly(combined_terminal, x = ~stock_price, color = ~scheme, type = "histogram",
        nbinsx = 60, opacity = 0.6, barmode = "overlay") |>
  layout(
    title = "Terminal Stock Price Distribution",
    xaxis = list(title = "S(T)"),
    yaxis = list(title = "Count")
  )
```

## Heston Characteristic Function Explorer

Plotly surfaces help interrogate the magnitude and real/imaginary parts across
time and frequency, echoing the lecture's characteristic-function focus.

```{r}
u_grid <- seq(-40, 40, length.out = 120)
tau_grid <- seq(0.25, 2, length.out = 30)

cf_data <- expand_grid(u = u_grid, maturity = tau_grid) |>
  mutate(
    value = heston_characteristic_function(
      u = u,
      maturity = maturity,
      initial_price = heston_base$initial_price,
      initial_variance = heston_base$initial_variance,
      risk_free_rate = heston_base$risk_free_rate,
      dividend_yield = heston_base$dividend_yield,
      mean_reversion = heston_base$mean_reversion,
      long_term_variance = heston_base$long_term_variance,
      vol_of_vol = heston_base$vol_of_vol,
      correlation = heston_base$correlation
    ),
    real_part = Re(value),
    imag_part = Im(value),
    magnitude = Mod(value)
  )

cf_real <- cf_data |>
  select(u, maturity, real_part) |>
  pivot_wider(names_from = maturity, values_from = real_part) |>
  arrange(u)

cf_imag <- cf_data |>
  select(u, maturity, imag_part) |>
  pivot_wider(names_from = maturity, values_from = imag_part) |>
  arrange(u)

cf_mag <- cf_data |>
  select(u, maturity, magnitude) |>
  pivot_wider(names_from = maturity, values_from = magnitude) |>
  arrange(u)

maturity_vals <- as.numeric(names(cf_real)[-1])
frequency_vals <- cf_real$u
```

```{r}
plot_ly(x = maturity_vals, y = frequency_vals, z = as.matrix(cf_real[, -1]),
        type = "surface", colorscale = "RdBu") |>
  layout(
    title = "Real Part of Heston Characteristic Function",
    scene = list(
      xaxis = list(title = "Maturity"),
      yaxis = list(title = "Frequency u"),
      zaxis = list(title = "Re(φ(u))")
    )
  )
```

```{r}
plot_ly(x = maturity_vals, y = frequency_vals, z = as.matrix(cf_imag[, -1]),
        type = "surface", colorscale = "RdBu") |>
  layout(
    title = "Imaginary Part of Heston Characteristic Function",
    scene = list(
      xaxis = list(title = "Maturity"),
      yaxis = list(title = "Frequency u"),
      zaxis = list(title = "Im(φ(u))")
    )
  )
```

```{r}
plot_ly(x = maturity_vals, y = frequency_vals, z = as.matrix(cf_mag[, -1]),
        type = "surface", colorscale = "Viridis") |>
  layout(
    title = "Magnitude of Heston Characteristic Function",
    scene = list(
      xaxis = list(title = "Maturity"),
      yaxis = list(title = "Frequency u"),
      zaxis = list(title = "|φ(u)|")
    )
  )
```

## Key Takeaways

- CIR variance paths expose boundary reflection under Euler and highlight the
  stability of AES sampling.
- AES preserves positivity more effectively, yielding smoother variance and
  stock surfaces.
- Monte Carlo call prices align with COS references once variance sampling is
  stabilised.
- Plotly surfaces provide an interactive view of the Heston characteristic
  function, mirroring the lecture's three-dimensional perspectives.
