---
title: "Implied Volatility Surfaces"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

`Lecture 04/ImpliedVolatility.py` walks through Newton-style solvers for
recovering implied volatilities from option quotes. The `CompFinanceR` package
wraps that workflow in tidy helpers while exposing confidence diagnostics.

### Learning Goals

- Construct call and put implied volatility surfaces.
- Validate solver convergence and bounds.
- Visualise smiles directly from tidy outputs.

## Setup

```{r}
library(CompFinanceR)
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
```

## Sample Option Data

We synthesise call prices and recover the implied volatility curve as in the
Python notebook.

```{r}
spot <- 100
risk_free <- 0.02
maturity <- 0.5
true_vol <- 0.25

call_spec <- black_scholes_spec(
  option_type = "call",
  strike = NA,
  maturity = maturity,
  risk_free_rate = risk_free,
  dividend_yield = 0
)

strikes <- seq(80, 120, by = 5)

calls <- tibble::tibble(
  strike = strikes,
  price = price_options(
    call_spec,
    spot = spot,
    volatility = true_vol
  )$price
)

calls
```

## Call Implied Volatility

```{r}
call_iv <- implied_volatility_call(
  option_data = calls,
  spot = spot,
  maturity = maturity,
  risk_free_rate = risk_free
)
call_iv
```

You can compare the recovered values with the input volatility to verify solver
accuracy.

```{r}
call_iv |>
  mutate(deviation = implied_volatility - true_vol)
```

## Put Implied Volatility

The helper `implied_volatility_put()` uses the same engine but flips the payoff
assumptions.

```{r}
puts <- calls |>
  mutate(price = price_options(
    black_scholes_spec(
      option_type = "put",
      strike = strike,
      maturity = maturity,
      risk_free_rate = risk_free
    ),
    spot = spot,
    volatility = true_vol
  )$price)

put_iv <- implied_volatility_put(
  option_data = puts,
  spot = spot,
  maturity = maturity,
  risk_free_rate = risk_free
)
```

## Visualising the Smile

```{r}
ggplot(call_iv, aes(strike, implied_volatility)) +
  geom_point(size = 2, colour = "steelblue") +
  geom_hline(yintercept = true_vol, linetype = "dashed") +
  labs(
    title = "Call Implied Volatility Curve",
    x = "Strike",
    y = "Implied Volatility"
  )
```

## Surface Construction

The lecture builds a grid across strikes and maturities. The `CompFinanceR`
functions fit naturally into `tidyr::expand_grid()` workflows.

```{r}
term_structure <- tidyr::expand_grid(
  strike = seq(80, 120, by = 5),
  maturity = c(0.25, 0.5, 1)
) |>
  mutate(price = price_options(
    black_scholes_spec(
      option_type = "call",
      strike = strike,
      maturity = maturity,
      risk_free_rate = risk_free
    ),
    spot = spot,
    volatility = true_vol
  )$price)


## Interactive 3D Surface

The Plotly view mirrors the lecture's surface visualisation while enabling
rotation and slicing along strike and maturity.

```{r}
surface <- term_structure |>
  distinct(strike, maturity, implied_volatility)

surface_wide <- surface_grid |>
  arrange(strike, maturity) |>
  tidyr::pivot_wider(names_from = maturity, values_from = implied_volatility)

maturity_vals <- names(surface_wide)[-1] |> as.numeric()

plot_ly(
  x = maturity_vals,
  y = surface_wide$strike,
  z = as.matrix(surface_wide[, -1]),
  type = "surface",
  colorscale = "Viridis"
) |>
  layout(
    scene = list(
      xaxis = list(title = "Maturity"),
      yaxis = list(title = "Strike"),
      zaxis = list(title = "Implied Volatility")
    ),
    title = "Interactive Implied Volatility Surface"
  )
```
  group_by(maturity) |>
  group_modify(~ implied_volatility_call(.x, spot = spot, maturity = unique(.x$maturity), risk_free_rate = risk_free))
```

```{r}
ggplot(surface, aes(strike, implied_volatility, colour = factor(maturity))) +
  geom_line() +
  labs(
    title = "Implied Volatility Surface Cuts",
    x = "Strike",
    y = "Implied Volatility",
    colour = "Maturity"
  )
```

## Diagnostics

The solver exposes how close the recovered volatility lies to the search
interval endpoints, mirroring the guard rails emphasised in the Python script.

```{r}
solver_diag <- implied_volatility_call(
  option_data = calls,
  spot = spot,
  maturity = maturity,
  risk_free_rate = risk_free,
  lower = 0.05,
  upper = 1
)
solver_diag
```

## Key Takeaways

- `implied_volatility_call()` and `implied_volatility_put()` encapsulate the
  Newton/bisection solver with guard rails discussed in Lecture 04.
- Tidy outputs make it easy to visualise smiles and surfaces.
- Diagnostics mirror the convergence checks implemented in the Python notebook.