---
title: "Jump Processes and the Merton Model"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

`Lecture 05/Jump Processes` explores Poisson processes and the Merton
jump-diffusion model via the scripts `PoissonProcess_paths.py` and
`MertonProcess_paths.py`. This vignette recreates those experiments in R.

### Learning Goals

- Simulate Poisson jump counts and event times.
- Generate Merton jump-diffusion price paths.
- Inspect how jump risk alters log-return distributions.

## Setup

```{r}
library(CompFinanceR)
library(dplyr)
library(ggplot2)
library(plotly)
set.seed(101)
```

## Poisson Process Simulation

`poisson_spec()` configures the intensity, while `simulate_paths()` produces the
cumulative jump count along a time grid.

```{r}
poisson <- poisson_spec(intensity = 0.8)

counts <- simulate_paths(
  poisson,
  n_paths = 5,
  n_steps = 100,
  maturity = 1
)
counts
```

```{r}
counts |>
  ggplot(aes(time, count, group = path_id)) +
  geom_step() +
  labs(title = "Poisson Jump Counts", y = "Count")
```

```{r}
plot_ly(counts, x = ~time, y = ~count, color = ~factor(path_id),
        type = "scatter", mode = "lines",
        line = list(shape = "hv"), showlegend = FALSE) |>
  layout(
    title = "Poisson Jump Counts (Interactive)",
    xaxis = list(title = "Time"),
    yaxis = list(title = "Count")
  )
```

## Merton Jump-Diffusion Paths

`merton_spec()` augments GBM with normally distributed jumps, matching the
parameters in the Python lecture.

```{r}
merton <- merton_spec(
  initial_price = 100,
  risk_free_rate = 0.02,
  volatility = 0.2,
  jump_intensity = 0.8,
  jump_mean = -0.05,
  jump_sd = 0.2
)

merton_paths <- simulate_paths(
  merton,
  n_paths = 5,
  n_steps = 250,
  maturity = 1,
  seed = 101
)
```

```{r}
merton_paths |>
  ggplot(aes(time, stock_price, group = path_id)) +
  geom_line() +
  labs(title = "Merton Jump-Diffusion", y = "Price")
```

```{r}
plot_ly(merton_paths, x = ~time, y = ~stock_price, color = ~factor(path_id),
        type = "scatter", mode = "lines", showlegend = FALSE) |>
  layout(
    title = "Merton Jump-Diffusion Paths (Interactive)",
    xaxis = list(title = "Time"),
    yaxis = list(title = "Price")
  )
```

## Comparing with Diffusion-Only GBM

```{r}
gbm_paths <- simulate_paths(
  gbm_spec(initial_value = 100, drift = 0.02, volatility = 0.2),
  n_paths = 10000,
  n_steps = 12,
  maturity = 1,
  seed = 101
)

merton_many <- simulate_paths(
  merton,
  n_paths = 10000,
  n_steps = 12,
  maturity = 1,
  seed = 101
)

summary_compare <- bind_rows(
  mutate(gbm_paths, model = "GBM"),
  mutate(merton_many, model = "Merton")
) |>
  filter(time == 1) |>
  summarise(
    mean_log_return = mean(log(stock_price / dplyr::first(stock_price))),
    sd_log_return = sd(log(stock_price / dplyr::first(stock_price))),
    .by = model
  )
summary_compare
```

Jump risk broadens the distribution and shifts the mean in line with the Python
insight.

## Histogram Comparison

```{r}
bind_rows(
  mutate(gbm_paths, model = "GBM"),
  mutate(merton_many, model = "Merton")
) |>
  filter(time == 1) |>
  mutate(log_return = log(stock_price / 100)) |>
  ggplot(aes(log_return, fill = model)) +
  geom_histogram(alpha = 0.6, bins = 60, position = "identity") +
  labs(title = "Terminal Log-Return Distribution")
```

```{r}
plot_ly(
  bind_rows(
    mutate(gbm_paths, model = "GBM"),
    mutate(merton_many, model = "Merton")
  ) |>
    filter(time == 1) |>
    mutate(log_return = log(stock_price / 100)),
  x = ~log_return,
  color = ~model,
  type = "histogram",
  nbinsx = 60,
  opacity = 0.6
) |>
  layout(
    title = "Terminal Log-Return Distribution (Interactive)",
    barmode = "overlay",
    xaxis = list(title = "Log return"),
    yaxis = list(title = "Count")
  )
```

## Key Takeaways

- `poisson_spec()` and `merton_spec()` encapsulate the jump process logic used
  in Lecture 05.
- The tidy simulation framework makes it easy to compare jump-diffusion against
  pure diffusion models.
- Jump-induced fat tails appear immediately in the log-return distribution,
  matching the Python demonstrations.