---
title: "Monte Carlo Diagnostics"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

`Lecture 09/Monte Carlo Simulation` analyses stochastic calculus basics,
quadratic variation, and numerical convergence through the scripts
`DeterministicFunction.py`, `StochasticIntegrals.py`, `dWdW.py`,
`EulerConvergence_GBM.py`, and `MilsteinConvergence_GBM.py`. This vignette
recreates those experiments with tidy helpers and interactive Plotly visuals.

### Learning Goals

- Compare hit-or-miss versus sample-mean Monte Carlo integration.
- Examine quadratic variation and stochastic integrals of Brownian motion.
- Measure Euler and Milstein convergence rates with interactive diagnostics.

## Setup

```{r}
library(CompFinanceR)
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
set.seed(2025)
```

## Monte Carlo Integration

We approximate \(\int_0^1 e^x dx\) using both hit-or-miss and sample-mean
estimators.

```{r}
exp_integrand <- function(x) exp(x)
true_value <- exp(1) - 1

hit_estimate <- mc_integrate_hit_or_miss(
  g = exp_integrand,
  lower = 0,
  upper = 1,
  lower_bound = 0,
  upper_bound = 3,
  n_samples = 20000,
  seed = 5
)

mean_estimate <- mc_integrate_sample_mean(
  g = exp_integrand,
  lower = 0,
  upper = 1,
  n_samples = 20000,
  seed = 5
)

integration_summary <- bind_rows(hit_estimate, mean_estimate) |>
  mutate(true_value = true_value, error = estimate - true_value)
integration_summary
```

### Hit-or-Miss Cloud

Interactive scatter replicates the lecture illustration of accepted and
rejected samples.

```{r}
mc_samples <- tibble::tibble(
  x = runif(4000, 0, 1),
  y = runif(4000, 0, 3)
) |>
  mutate(g_of_x = exp_integrand(x), hit = y <= g_of_x)

curve_points <- tibble::tibble(
  x = seq(0, 1, length.out = 200),
  g_of_x = exp_integrand(x)
)

plot_ly(mc_samples, x = ~x, y = ~y, color = ~hit, colors = c("firebrick", "seagreen"),
        type = "scatter", mode = "markers",
        marker = list(size = 4, opacity = 0.55)) |>
  add_lines(data = curve_points, x = ~x, y = ~g_of_x, name = "exp(x)",
            line = list(color = "steelblue", width = 3)) |>
  layout(
    title = "Hit-or-Miss Monte Carlo Integration",
    xaxis = list(title = "x"),
    yaxis = list(title = "y"),
    legend = list(title = list(text = "Sample Accepted"))
  )
```

## Quadratic Variation Diagnostics

`dWdW.py` studies \(E[(\nabla W)^2]\) and its variance. The helper
`quadratic_variation_diagnostics()` reproduces that behaviour.

```{r}
qv_results <- quadratic_variation_diagnostics(
  maturity = 1,
  n_paths = 20000,
  step_grid = seq(10, 200, by = 10),
  seed = 9
)
qv_results |> slice_tail(n = 5)
```

```{r}
plot_ly() |>
  add_lines(data = qv_results, x = ~dt, y = ~mean_sq_increment,
            name = "Empirical Mean", line = list(color = "steelblue")) |>
  add_lines(data = qv_results, x = ~dt, y = ~expected_mean,
            name = "Theoretical Mean", line = list(color = "black", dash = "dash")) |>
  add_lines(data = qv_results, x = ~dt, y = ~var_sq_increment,
            name = "Empirical Variance", yaxis = "y2",
            line = list(color = "tomato")) |>
  add_lines(data = qv_results, x = ~dt, y = ~expected_variance,
            name = "Theoretical Variance", yaxis = "y2",
            line = list(color = "black", dash = "dot")) |>
  layout(
    title = "Quadratic Variation Diagnostics",
    xaxis = list(title = "dt"),
    yaxis = list(title = "Mean of (ΔW)^2"),
    yaxis2 = list(overlaying = "y", side = "right", title = "Variance of (ΔW)^2"),
    legend = list(orientation = "h")
  )
```

## Brownian Motion Integrals

`StochasticIntegrals.py` tracks Brownian motion and the associated Riemann and
Itô integrals. The `simulate_stochastic_integrals()` helper returns all three in
one tidy tibble.

```{r}
integrals <- simulate_stochastic_integrals(
  maturity = 1,
  n_paths = 1,
  n_steps = 1500,
  seed = 17
)
one_path <- integrals |>
  filter(path_id == 1)
```

```{r}
plot_ly(one_path, x = ~time, y = ~brownian, name = "W(t)", type = "scatter", mode = "lines",
        line = list(color = "royalblue")) |>
  add_lines(y = ~riemann_integral, name = "∫ W(s) ds", line = list(color = "firebrick")) |>
  add_lines(y = ~ito_integral, name = "∫ W(s) dW(s)", line = list(color = "forestgreen")) |>
  layout(
    title = "Brownian Motion and Its Integrals",
    xaxis = list(title = "Time"),
    yaxis = list(title = "Value"),
    legend = list(orientation = "h")
  )
```

The terminal Itô integral matches \((W_T^2 - T)/2\) up to simulation error.

```{r}
terminal_check <- one_path |>
  slice_tail(n = 1) |>
  mutate(ito_formula = (brownian^2 - 1) / 2,
         difference = ito_integral - ito_formula)
terminal_check
```

For a broader view we can inspect a surface of Brownian paths reminiscent of
the lecture animations.

```{r}
brownian_surface <- simulate_stochastic_integrals(
  maturity = 1,
  n_paths = 25,
  n_steps = 300,
  seed = 42
) |>
  select(path_id, time, brownian) |>
  pivot_wider(names_from = time, values_from = brownian) |>
  arrange(path_id)

time_vals <- as.numeric(names(brownian_surface)[-1])
path_matrix <- as.matrix(brownian_surface[, -1])

plot_ly(x = time_vals, y = brownian_surface$path_id, z = path_matrix,
        type = "surface", colorscale = "GnBu") |>
  layout(
    title = "Brownian Motion Surface",
    scene = list(
      xaxis = list(title = "Time"),
      yaxis = list(title = "Path"),
      zaxis = list(title = "W(t)")
    )
  )
```

## Euler vs Milstein Convergence

`EulerConvergence_GBM.py` and `MilsteinConvergence_GBM.py` benchmark the weak
error across discretisations. We replicate that analysis using
`euler_convergence_study()` and `milstein_convergence_study()`.

```{r}
step_grid <- c(4, 8, 16, 32, 64, 128, 256)
convergence <- bind_rows(
  euler_convergence_study(
    initial_price = 100,
    drift = 0.05,
    volatility = 0.3,
    maturity = 1,
    n_paths = 20000,
    step_grid = step_grid,
    seed = 7
  ),
  milstein_convergence_study(
    initial_price = 100,
    drift = 0.05,
    volatility = 0.3,
    maturity = 1,
    n_paths = 20000,
    step_grid = step_grid,
    seed = 7
  )
)
convergence
```

```{r}
plot_ly(convergence, x = ~log10(dt), y = ~log10(rmse), color = ~scheme,
        type = "scatter", mode = "lines+markers",
        colors = c("#1f77b4", "#ff7f0e")) |>
  layout(
    title = "Euler vs Milstein Convergence (log-log)",
    xaxis = list(title = "log10(dt)"),
    yaxis = list(title = "log10(RMSE)"),
    legend = list(x = 0.02, y = 0.98)
  )
```

Add a reference slope of one to match the lecture's log-log diagnostic.

```{r}
anchor <- convergence |>
  arrange(dt) |>
  slice(1) |>
  summarise(dt = dt, rmse = rmse)
reference <- tibble::tibble(
  log_dt = log10(range(convergence$dt)),
  log_rmse = log10(anchor$rmse) + 1 * (log_dt - log10(anchor$dt))
)

plot_ly(convergence, x = ~log10(dt), y = ~log10(rmse), color = ~scheme,
        type = "scatter", mode = "lines+markers") |>
  add_lines(data = reference, x = ~log_dt, y = ~log_rmse,
            name = "Reference slope 1", line = list(color = "gray40", dash = "dash")) |>
  layout(
    title = "Weak Convergence Rate Comparison",
    xaxis = list(title = "log10(dt)"),
    yaxis = list(title = "log10(RMSE)")
  )
```

## Key Takeaways

- Tidy Monte Carlo helpers mirror the lecture's integration, stochastic
  calculus, and convergence diagnostics.
- Plotly reproduces the interactive figures (hit-or-miss clouds, Brownian
  surfaces, log-log convergence) used in the Python notebooks.
- Euler and Milstein schemes follow the expected slope-1 weak convergence when
  visualised on log-log axes, confirming the lecture conclusions.
