---
title: "Pathwise Greeks via Monte Carlo"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

`Lecture 11/Hedging and Monte Carlo Sensitivities/PathwiseSens_DeltaVega.py`
illustrates the pathwise differentiation method for computing delta and vega.
This vignette mirrors that workflow with the `CompFinanceR` pathwise utilities
and augments the analysis with interactive Plotly visualisations.

### Learning Goals

- Estimate delta and vega using the pathwise method under geometric Brownian
  motion.
- Compare Monte Carlo Greeks with finite-difference and analytical benchmarks.
- Inspect estimator convergence and the contribution distribution interactively.

## Setup

```{r}
library(CompFinanceR)
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
set.seed(451)
```

We consider a call option with the same parameters used in the lecture notes.

```{r}
option_spec <- gbm_spec(
  initial_value = 1,
  drift = 0.06,
  volatility = 0.3
)
strike <- 1
maturity <- 1
risk_free <- 0.06
```

## Pathwise vs Analytical Benchmarks

Using a large Monte Carlo sample we compare the pathwise method with central
finite differences and Black-Scholes analytical Greeks.

```{r}
comparison <- compare_with_finite_diff(
  process_spec = option_spec,
  strikes = strike,
  maturity = maturity,
  n_paths = 20000,
  n_steps = 500,
  risk_free_rate = risk_free,
  option_type = "call",
  bump_spot = 0.01,
  bump_vol = 5e-4,
  seed = 123
)
comparison
```

## Convergence with Path Count

We reproduce the lecture figure showing pathwise estimators converging towards
the analytical values as the number of paths increases.

```{r}
path_grid <- unique(round(seq(200, 8000, length.out = 20)))

convergence <- purrr::map_dfr(
  path_grid,
  \(n_paths) {
    paths <- simulate_paths(
      process_spec = option_spec,
      n_paths = n_paths,
      n_steps = 400,
      maturity = maturity,
      seed = 100
    )
    tibble::tibble(
      n_paths = n_paths,
      delta = pathwise_delta(
        paths,
        strikes = strike,
        maturity = maturity,
        risk_free_rate = risk_free,
        option_type = "call",
        initial_price = option_spec$initial_value
      )$delta_pathwise,
      vega = pathwise_vega(
        paths,
        strikes = strike,
        maturity = maturity,
        risk_free_rate = risk_free,
        option_type = "call",
        volatility = option_spec$volatility,
        initial_price = option_spec$initial_value
      )$vega_pathwise
    )
  }
)

analytic_greeks <- black_scholes_spec(
  option_type = "call",
  strike = strike,
  maturity = maturity,
  risk_free_rate = risk_free
) |>
  price_options(spot = option_spec$initial_value, volatility = option_spec$volatility) |>
  select(delta_analytic = delta, vega_analytic = vega)
analytic_greeks
```

```{r}
plot_ly(convergence, x = ~n_paths, y = ~delta, type = "scatter",
        mode = "lines+markers", name = "Delta (pathwise)",
        line = list(color = "#1f77b4")) |>
  add_lines(x = ~n_paths, y = analytic_greeks$delta_analytic,
            name = "Delta (analytic)", line = list(color = "black", dash = "dash")) |>
  layout(
    title = "Pathwise Delta Convergence",
    xaxis = list(title = "Number of paths"),
    yaxis = list(title = "Delta")
  )
```

```{r}
plot_ly(convergence, x = ~n_paths, y = ~vega, type = "scatter",
        mode = "lines+markers", name = "Vega (pathwise)",
        line = list(color = "#ff7f0e")) |>
  add_lines(x = ~n_paths, y = analytic_greeks$vega_analytic,
            name = "Vega (analytic)", line = list(color = "black", dash = "dash")) |>
  layout(
    title = "Pathwise Vega Convergence",
    xaxis = list(title = "Number of paths"),
    yaxis = list(title = "Vega")
  )
```

## Contribution Distributions

The lecture emphasises how each path contributes to the estimator. We reproduce
the contribution histograms interactively.

```{r}
large_paths <- simulate_paths(
  process_spec = option_spec,
  n_paths = 20000,
  n_steps = 400,
  maturity = maturity,
  seed = 321
) |>
  filter(time == maturity)

delta_contrib <- large_paths |>
  mutate(
    payoff_indicator = as.numeric(stock_price > strike),
    delta_contribution = exp(-risk_free * maturity) * (stock_price / option_spec$initial_value) * payoff_indicator
  )

vega_contrib <- large_paths |>
  mutate(
    payoff_indicator = as.numeric(stock_price > strike),
    adjustment = (log(stock_price / option_spec$initial_value) - (risk_free + 0.5 * option_spec$volatility^2) * maturity),
    vega_contribution = exp(-risk_free * maturity) * (stock_price / option_spec$volatility) * adjustment * payoff_indicator
  )
```

```{r}
plot_ly(delta_contrib, x = ~delta_contribution, type = "histogram", nbinsx = 60,
        marker = list(color = "steelblue")) |>
  layout(
    title = "Distribution of Pathwise Delta Contributions",
    xaxis = list(title = "Contribution"),
    yaxis = list(title = "Count")
  )
```

```{r}
plot_ly(vega_contrib, x = ~vega_contribution, type = "histogram", nbinsx = 60,
        marker = list(color = "darkorange")) |>
  layout(
    title = "Distribution of Pathwise Vega Contributions",
    xaxis = list(title = "Contribution"),
    yaxis = list(title = "Count")
  )
```

## Hedging Context

Pathwise Greeks plug directly into the hedging routines developed elsewhere in
the course. For completeness we show the Monte Carlo delta and vega alongside
finite differences on a tidy grid of strikes.

```{r}
strike_grid <- seq(0.6, 1.4, by = 0.1)

comparison_grid <- compare_with_finite_diff(
  process_spec = option_spec,
  strikes = strike_grid,
  maturity = maturity,
  n_paths = 15000,
  n_steps = 300,
  risk_free_rate = risk_free,
  option_type = "call",
  bump_spot = 0.01,
  bump_vol = 5e-4,
  seed = 777
)
comparison_grid
```

Interactive scatter plots help gauge deviations strike by strike.

```{r}
comparison_long <- comparison_grid |>
  pivot_longer(
    cols = c(delta_pathwise, delta_fd, delta_analytic),
    names_to = "metric",
    values_to = "value"
  )

plot_ly(comparison_long, x = ~strike, y = ~value, color = ~metric,
        type = "scatter", mode = "lines+markers") |>
  layout(
    title = "Strike-wise Delta Comparison",
    xaxis = list(title = "Strike"),
    yaxis = list(title = "Delta")
  )
```

```{r}
vega_long <- comparison_grid |>
  pivot_longer(
    cols = c(vega_pathwise, vega_fd, vega_analytic),
    names_to = "metric",
    values_to = "value"
  )

plot_ly(vega_long, x = ~strike, y = ~value, color = ~metric,
        type = "scatter", mode = "lines+markers") |>
  layout(
    title = "Strike-wise Vega Comparison",
    xaxis = list(title = "Strike"),
    yaxis = list(title = "Vega")
  )
```

## Key Takeaways

- Pathwise estimators converge rapidly to the analytical Greeks, matching the
  lecture demonstrations.
- Interactive histograms reveal the contribution distributions that underpin
  estimator variance.
  
- Finite-difference and pathwise estimates agree across strikes, validating the
  R implementation against the Python baseline.
