---
title: "Simulating GBM and ABM Paths"
format:
  html:
    toc: true
    toc-location: left
execute:
  message: false
  warning: false
  echo: true
---

## Motivation

`Lecture 03/Option Pricing and Simulation in Python` introduces Brownian-driven
models under both arithmetic and geometric dynamics and contrasts physical and
risk-neutral measures. This vignette reproduces the Python scripts
`GBM_ABM_paths.py`, `GBM_ABM_paths_Martingale.py`, and
`PathsUnderQandPmeasure.py` using the unified R API.

### Learning Goals

- Simulate arithmetic and geometric Brownian motion in tidy format.
- Verify martingale properties under the risk-neutral measure.
- Compare drift adjustments between real-world (P) and risk-neutral (Q)
  dynamics.

## Package Setup

```{r}
library(CompFinanceR)
library(dplyr)
library(ggplot2)
library(plotly)
set.seed(123)
```

## Arithmetic Brownian Motion

`abm_spec()` defines the drift and volatility for arithmetic Brownian motion.

```{r}
abm <- abm_spec(
  initial_value = 100,
  drift = 0.05,
  volatility = 0.2
)

abm_paths <- simulate_paths(
  process_spec = abm,
  n_paths = 5,
  n_steps = 252,
  maturity = 1
)

abm_paths
```

A quick plot mirrors the Python notebook output.

```{r}
abm_paths |>
  ggplot(aes(time, stock_price, group = path_id)) +
  geom_line(alpha = 0.7) +
  labs(title = "Arithmetic Brownian Motion", x = "Time", y = "Level")
```

```{r}
plot_ly(abm_paths, x = ~time, y = ~stock_price, color = ~factor(path_id),
        type = "scatter", mode = "lines", showlegend = FALSE,
        hovertemplate = paste(
          "Path %{color}<br>",
          "t = %{x:.3f}<br>",
          "X_t = %{y:.3f}",
          sep = ""
        )) |>
  layout(
    title = "ABM Paths (Interactive)",
    xaxis = list(title = "Time"),
    yaxis = list(title = "Level")
  )
```

## Geometric Brownian Motion

`gbm_spec()` matches the familiar Black-Scholes dynamics.

```{r}
gbm <- gbm_spec(
  initial_value = 100,
  drift = 0.05,
  volatility = 0.2
)

gbm_paths <- simulate_paths(gbm, n_paths = 5, n_steps = 252, maturity = 1)
```

The simulated paths stay positive by construction, as in the lecture figures.

```{r}
gbm_paths |>
  ggplot(aes(time, stock_price, group = path_id)) +
  geom_line(alpha = 0.7) +
  labs(title = "Geometric Brownian Motion", x = "Time", y = "Price")
```

```{r}
plot_ly(gbm_paths, x = ~time, y = ~stock_price, color = ~factor(path_id),
        type = "scatter", mode = "lines", showlegend = FALSE,
        hovertemplate = paste(
          "Path %{color}<br>",
          "t = %{x:.3f}<br>",
          "S_t = %{y:.3f}",
          sep = ""
        )) |>
  layout(
    title = "GBM Paths (Interactive)",
    xaxis = list(title = "Time"),
    yaxis = list(title = "Price")
  )
```

## Martingale Check Under Q

The Python script shows that discounted GBM under the risk-neutral measure is a
martingale. We recover the same result by setting the drift equal to the
risk-free rate.

```{r}
risk_free <- 0.02
q_spec <- gbm_spec(initial_value = 100, drift = risk_free, volatility = 0.2)

q_paths <- simulate_paths(q_spec, n_paths = 1000, n_steps = 252, maturity = 1)

martingale <- q_paths |>
  mutate(discounted_price = stock_price * exp(-risk_free * time)) |>
  group_by(time) |>
  summarise(mean_discounted = mean(discounted_price), .groups = "drop")

martingale
```

```{r}
martingale |>
  ggplot(aes(time, mean_discounted)) +
  geom_line() +
  geom_hline(yintercept = q_spec$initial_value, linetype = "dashed") +
  labs(
    title = "Discounted GBM Under Q",
    x = "Time",
    y = "Mean Discounted Price"
  )
```

```{r}
plot_ly(martingale, x = ~time, y = ~mean_discounted, type = "scatter",
        mode = "lines", line = list(color = "steelblue"),
        name = "E[e^{-rt} S_t]") |>
  add_lines(x = ~time, y = q_spec$initial_value, name = "S0",
            line = list(color = "black", dash = "dash"), inherit = FALSE) |>
  layout(
    title = "Discounted GBM Under Q (Interactive)",
    xaxis = list(title = "Time"),
    yaxis = list(title = "Mean Discounted Price")
  )
```

The series stays near the initial price, matching the martingale property
illustrated by `GBM_ABM_paths_Martingale.py`.

## Comparing P and Q Measures

`PathsUnderQandPmeasure.py` contrasts the drift adjustment required under the
risk-neutral measure. We can recreate the same comparison by simulating under
both drifts and examining expectations.

```{r}
paths_compare <- bind_rows(
  mutate(simulate_paths(gbm, n_paths = 2000, n_steps = 252, maturity = 1), measure = "P"),
  mutate(simulate_paths(q_spec, n_paths = 2000, n_steps = 252, maturity = 1), measure = "Q")
)

terminal_summary <- paths_compare |>
  filter(time == 1) |>
  group_by(measure) |>
  summarise(
    mean_price = mean(stock_price),
    sd_price = sd(stock_price),
    .groups = "drop"
  )
terminal_summary
```

The risk-neutral measure pushes the expected terminal price to match the
forward price `S0 * exp((r - q) * T)` while preserving the diffusion term,
which is the key conceptual takeaway from the Python exercises.

## Key Takeaways

- `simulate_paths()` unifies ABM and GBM experiments with minimal boilerplate.
- Discounting under the risk-neutral measure produces the martingale check that
  features prominently in Lecture 03.
- Switching between real-world and risk-neutral drifts is as simple as changing
  the `drift` parameter on the specification.